---
title: "Test EpiNow2"
output: html_notebook
---

## Test EpiNow2 on SEIR synthetic data
Katie Gostic
26 June 2020



## Install packages
```{r}
# install.packages("drat")
# drat:::add("epiforecasts")
# install.packages("EpiNow2")
```
## Load packages
```{r}
rm(list = ls())
setwd(here::here())
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(EpiNow2)
library(NCoVUtils)
library(deSolve)
require(data.table, quietly = TRUE) 
require(future, quietly = TRUE)
require(forecastHybrid, quietly = TRUE)
ggplot2::theme_set(theme_bw())
```


## Generate synthetic data
```{r}
source('util.R')
source('simulation.R')

## Set simulation parameters
parlist <- {
  list(
    N = 2e6, #total population size
    E_init = 0, # initial in E class
    I_init = 60, # initial in I class
    t_E = 4, # mean time in E (latent period)
    t_I = 4, # mean time in I (duration of infectiousness)
    n_t = 300, # total timesteps
    R0_vals = c(2, 0.7, 1.1), ## Steady-state R0 values held at different points in the simulation
    change_starts = c(50, 90), ## Times at which R0 starts to decrease from its previous steady state value
    change_ends = c(57, 97), ## Times at which R0 reaches its new steady state
    model_types = c('seir'), # Can also choose sir
    methods = c('ode', 'stochastic') # could also choose ode
  )
}

## Derive the mean and variance of the serial interval from the input parameters
parlist$true_mean_GI = (parlist$t_E+parlist$t_I)
parlist$true_var_GI = 2*(parlist$true_mean_GI/2)^2



## Simulate SEIR data using a stochastic (ode) model
## Results are saved to subdirectory rds/
arnaught <- with(parlist, specify_arnaught(R0_vals, change_starts, change_ends, n_t))
sim_wrapper(arnaught, parlist)


write_rds(parlist, path = 'true_pars.rds')
```

## Visualize synthetic data
```{r}
get_sim_df() %>%
    filter(time < 300) %>%
ggplot() +
  geom_line(aes(x = time, y = infections))+
  geom_vline(aes(xintercept = parlist$change_starts[1]), lty = 2)+ ## Dahsed line where Rt starts to decrease
  geom_vline(aes(xintercept = parlist$change_starts[2]), lty = 2)+
  ggtitle('Epidemic curve') -> inc

get_sim_df() %>% 
  filter(time < 300) %>%
  ggplot()+
  geom_line(aes(x = time, y = true_rt)) +
  geom_hline(aes(yintercept = 1), lty = 2)+
  ylab(expression(paste(R[t])))+
  ggtitle(expression(paste('Underlying ', R[t], ' values'))) -> R0

cowplot::plot_grid(R0, inc,  nrow = 2)
ggsave('figs/SEIR_sim.png', width = 5, height = 5, units = 'in', dpi = 300)
```

## Set delays, generation interval and incubation period for input into EpiNow2
```{r}
xx = seq(0, 30, by = 0.01)
gi_samples = rgamma(1000, get_shape(parlist$true_mean_GI, parlist$true_var_GI), get_rate(parlist$true_mean_GI, parlist$true_var_GI))
## Fit the true generation time to a lognormal
lnormal_gi_pars <- optim(par = c(mm = log(8), sd = 1), fn = function(pars){
  -sum(dlnorm(gi_samples, pars['mm'], pars['sd'], log = TRUE))
})


generation_time <- list(mean = lnormal_gi_pars$par[1],
                        mean_sd = 1,
                        sd = lnormal_gi_pars$par[2],
                        sd_sd = 1,
                        max = 30)
{
plot(xx, dlnorm(xx, generation_time$mean, generation_time$sd), type = 'l', main = 'assumed generation time', xlab = 'days', ylab = 'dens')
lines(xx, 
      dgamma(xx, get_shape(parlist$true_mean_GI, parlist$true_var_GI), 
             get_rate(parlist$true_mean_GI, parlist$true_var_GI)), 
             col = 'blue')
}

incubation_period <- list(mean = EpiNow2::covid_incubation_period[1, ]$mean,
                          mean_sd = EpiNow2::covid_incubation_period[1, ]$mean_sd,
                          sd = EpiNow2::covid_incubation_period[1, ]$sd,
                          sd_sd = EpiNow2::covid_incubation_period[1, ]$sd_sd,
                          max = 30)
plot(xx, dlnorm(xx, incubation_period$mean, incubation_period$sd), type = 'l', main = 'assumed incubation time', xlab = 'days', ylab = 'dens')

obs_rep_delay <- list(mean = log(5), ## assume mean of 5 days for outpatient testing (48h from symptom onset + 48h for reporting + report the following day)
                         mean_sd = 1,
                         sd = log(1.7),
                         sd_sd = 1,
                         max = 30)
plot(xx, dlnorm(xx, log(5), log(1.7)), type = 'l', main = 'assumed delay from onset to case detection', xlab = 'days', ylab = 'dens')

## For death, assume a mean of 15 days from symptom onset and an sd of about 7d
death_rep_delay <- list(mean = log(15), ## mean of 3 days
                         mean_sd = 1,
                         sd = log(1.5),
                         sd_sd = 1,
                         max = 30)
plot(xx, dlnorm(xx, log(15), log(1.5)), type = 'l', main = 'assumed delay from onset to death', xlab = 'days', ylab = 'dens')
```

## Use the same delays to generate an observed time series from the synthetic data
```{r}
## Function to draw random samples from the incubation period distribution
r_inc_dist <- function(n){rlnorm(n, meanlog = incubation_period$mean, sdlog = incubation_period$sd)} 


## Function to draw samples from delays to observation:
outpatient_delay_dist <- function(nn, inc_dist = r_inc_dist){
  r_sym_to_obs_dist <- function(n){rlnorm(n, obs_rep_delay$mean, obs_rep_delay$sd)} # Additional delay from symptoms -> observation
  inc_dist(nn) + r_sym_to_obs_dist(nn)
}

## Set the delay distribution to death
death_delay_dist <- function(nn, inc_dist = r_inc_dist){
  r_sym_to_obs_dist <- function(n){rlnorm(n, death_rep_delay$mean, death_rep_delay$sd)} # Additional delay from symptoms -> observation
  inc_dist(nn) + r_sym_to_obs_dist(nn)
}
```

## Generate synthetic time series of observations from infections
```{r}
source('inf_to_obs.R')

## Wrapper function to get synthetic data in a data frame that includes times of observation
get_obs_ts <- function(mt = NULL){ # "Max time" - time at which to end the time series of observations
  if(length(mt) == 0){mt = max(get_sim_df()$time)}
  
  ## Sample from delays to get times of observation
  get_sim_df() %>%
    merge(
      get_tObs_from_tInf(get_sim_df()$incidence, get_sim_df()$time, outpatient_delay_dist, return_times = T),
      by = 'time', all = TRUE) %>% rename(obs_outpatient = n) %>%
    merge(
      get_tObs_from_tInf(get_sim_df()$incidence, get_sim_df()$time, death_delay_dist, return_times = T),
      by = 'time', all = TRUE) %>% rename(obs_deaths = n) %>%
    as.tbl() -> obs_df
  
  ## Truncate at time mt
  obs_df <- obs_df %>% filter(time <= mt)
  return(obs_df)
}

## Generate synthetic times of observation up to t = 150
obs_df <- get_obs_ts(150) %>%
    mutate(date = Sys.Date()-rev(time)) ## Add calendar dates for input into epinow

## Plot
obs_df %>%
  select(time, incidence, contains('obs')) %>%
  pivot_longer(-time, names_to = 'data_type', values_to = 'count') %>%
  mutate(is_imputed = ifelse(!grepl(pattern = 'obs', x = data_type), 'observed', 'underlying'),
         data_type = factor(data_type, 
                            levels = c('incidence', 'obs_outpatient', 'imputed_hospital', 'obs_deaths'),
                            labels = c('latent (true) infections\nlater observed as...', 'cases', 'hospital admissions', 'deaths'))) %>%
  ggplot() +
  geom_line(aes(x = time, y = count, color = data_type, lty = is_imputed)) +
  ggtitle('Synthetic Data') +
  scale_color_viridis_d(direction = -1)+
  guides(color = 'none')+
  theme(legend.position = 'bottom')+
  labs(color = '', linetype = 'Type') -> obs
obs + guides(color = 'legend') + theme(legend.position = 'right')
```


## Input into epinow2 with everything perfectly specified
```{r}
## Write a wrapper to reformat the desired synthetic data for input into epiEstim
get_in_dat <- function(obs_colname, obs_df = obs_df){
  obs_df[,c('date', obs_colname)] %>%
    setnames(c('date', 'confirm')) %>%
    as.data.table() %>%
    return()
}
```

## Test using synthetic case observations
```{r}
## From outpatient testing
est_from_cases <- EpiNow2::epinow(reported_cases = get_in_dat('obs_outpatient'), 
                             generation_time = generation_time,  ## Generation time priors
                             incubation_period = incubation_period, ## Incubation priors
                             reporting_delay = obs_rep_delay, ## Reporting dealy priors
                             rt_prior = list(mean = 1, sd = 1), horizon = 7,
                             samples = 2000, warmup = 500, cores = 4,
                             chains = 4, verbose = TRUE,
                             target_folder = 'perfectly_specified')
```

## Test using synthetic deaths
```{r}
est_from_deaths <- EpiNow2::epinow(reported_cases = get_in_dat('obs_deaths'), 
                             generation_time = generation_time,  ## Generation time priors
                             incubation_period = incubation_period, ## Incubation priors
                             reporting_delay = death_rep_delay, ## Reporting dealy priors
                             rt_prior = list(mean = 1, sd = 1), horizon = 7,
                             samples = 2000, warmup = 500, cores = 4,
                             chains = 4, verbose = TRUE,
                             target_folder = 'perfectly_specified')
```

## Repeat for a truncated time series
```{r}
## Generate synthetic times of observation up to t = 150
trunc_df <- get_obs_ts(97) %>%
    mutate(date = Sys.Date()-rev(time)) ## Add calendar dates for input into epinow

## Plot
trunc_df %>%
  select(time, incidence, contains('obs')) %>%
  pivot_longer(-time, names_to = 'data_type', values_to = 'count') %>%
  mutate(is_imputed = ifelse(!grepl(pattern = 'obs', x = data_type), 'observed', 'underlying'),
         data_type = factor(data_type, 
                            levels = c('incidence', 'obs_outpatient', 'imputed_hospital', 'obs_deaths'),
                            labels = c('latent (true) infections\nlater observed as...', 'cases', 'hospital admissions', 'deaths'))) %>%
  ggplot() +
  geom_line(aes(x = time, y = count, color = data_type, lty = is_imputed)) +
  ggtitle('Synthetic Data') +
  scale_color_viridis_d(direction = -1)+
  guides(color = 'none')+
  theme(legend.position = 'bottom')+
  labs(color = '', linetype = 'Type') -> obs
obs + guides(color = 'legend') + theme(legend.position = 'right')
```


## Test using synthetic cases
```{r}
est_from_cases_t <- EpiNow2::epinow(reported_cases = get_in_dat('obs_outpatient', trunc_df), 
                             generation_time = generation_time,  ## Generation time priors
                             incubation_period = incubation_period, ## Incubation priors
                             reporting_delay = obs_rep_delay, ## Reporting dealy priors
                             rt_prior = list(mean = 1, sd = 1), horizon = 7,
                             samples = 2000, warmup = 500, cores = 4,
                             chains = 4, verbose = TRUE,
                             target_folder = 'truncated_94')
```

## Test using synthetic deaths
```{r}
est_from_deaths_t <- EpiNow2::epinow(reported_cases = get_in_dat('obs_deaths', trunc_df), 
                             generation_time = generation_time,  ## Generation time priors
                             incubation_period = incubation_period, ## Incubation priors
                             reporting_delay = death_rep_delay, ## Reporting dealy priors
                             rt_prior = list(mean = 1, sd = 1), horizon = 7,
                             samples = 2000, warmup = 500, cores = 4,
                             chains = 4, verbose = TRUE,
                             target_folder = 'truncated_94')
```





## Plot posteriors vs. priors
```{r}
plot_pr_vs_post <- function(
  ests,
  incubation_period,
  generation_time,
  delay_time
){
  gi_prior <- data_frame(
    xx = seq(0, generation_time$max, by = 0.01),
    dens = dlnorm(xx, generation_time$mean, generation_time$sd)
  )
  inc_prior <- data_frame(
    xx = seq(0, incubation_period$max, by = 0.01),
    dens = dlnorm(xx, incubation_period$mean, incubation_period$sd)
  )
  del_prior <- data_frame(
    xx = seq(0, delay_time$max, by = 0.01),
    dens = dlnorm(xx, delay_time$mean, delay_time$sd)
  )
  
  ## Write a function to evaluate the posterior distribution at each timestep
  get_post <- function(ests, truepars, vname = 'gen_interval'){
    ests$estimates$samples %>%
      as.tbl() %>%
      filter(grepl(pattern = 'gt', variable)) %>%
      group_by(variable, sample) %>%
      summarise(value = unique(value)) %>%
      pivot_wider(names_from = variable, values_from = value) %>%
      apply(MARGIN = 1, FUN = function(ss) dlnorm(seq(0, truepars$max, by = 0.01), ss[2], ss[3])) %>%
      apply(MARGIN = 1, function(rr){
        c(lower = quantile(rr, .025), 
          median = quantile(rr, .5),
          upper = quantile(rr, .975),
          mean = mean(rr))
      }) %>%
      t() %>%
      as.data.frame() %>%
      mutate(xx = seq(0, truepars$max, by = 0.01),
             variable = vname)
  }
  
  gi_post <- get_post(ests, truepars = generation_time, vname = 'gi_post')
  inc_post <- get_post(ests, truepars = incubation_period, vname = 'inc_post')
  delay_post <- get_post(ests, truepars = delay_time, vname = 'delay_post')
  
  gg <- ggplot()+
    geom_ribbon(data = gi_post, aes(x = xx, ymin = `lower.2.5%`, ymax = `upper.97.5%`), fill = 'yellow', alpha = .5)+
    geom_line(data = gi_post, aes(x = xx, y = mean), color = 'yellow', alpha = .5)+
    geom_line(data = gi_prior, aes(x = xx, y = dens), lty = 2) +
    ggtitle('Generation interval')
  
  ii <- ggplot()+
    geom_ribbon(data = inc_post, aes(x = xx, ymin = `lower.2.5%`, ymax = `upper.97.5%`), fill = 'yellow', alpha = .5)+
    geom_line(data = inc_post, aes(x = xx, y = mean), color = 'yellow', alpha = .5)+
    geom_line(data = inc_prior, aes(x = xx, y = dens), lty = 2)+
    ggtitle('Incubation period')
  
  dd <- ggplot()+
    geom_ribbon(data = delay_post, aes(x = xx, ymin = `lower.2.5%`, ymax = `upper.97.5%`), fill = 'yellow', alpha = .5)+
    geom_line(data = delay_post, aes(x = xx, y = mean), color = 'yellow', alpha = .5)+
    geom_line(data = del_prior, aes(x = xx, y = dens), lty = 2)+
    ggtitle('Reporting delay')
  
  cowplot::plot_grid(gg, ii, dd, ncol = 3)
}
```
