---
title: "Test shift for infections prior"
output: html_notebook
---

## Setup
```{r}
knitr::opts_chunk$set(echo = TRUE)
## Estimate Rt using epinow2
## Test if the assumed delay to case or death observation is three days too long and much more variable than the true delay

## Load dependencies and set parameters ------------------------------------------------------
rm(list = ls())
source('../00-load_packages.R')
source('../00-util.R')
source('../00-run_test.R')
ggplot2::theme_set(theme_bw())

## Check if synthetic data already exists.
## If so, load from cache
## If not, make the data
parlist <- load_parlist()
# Synthetic data is loaded using:
# get_sim_df()

## Set parameters for EpiNow2 test
testpars <- list(
  last_obs_time = 150,
  output_folder = 'perfectly_specified',
  ## True delays are used to generate data
  true_log_mean_case_delay = log(5),
  true_log_sd_case_delay = log(1.7),
  true_log_mean_death_delay = log(15),
  true_log_sd_death_delay = log(1.5),
  true_log_mean_inc = EpiNow2::covid_incubation_period[1, ]$mean,
  true_log_sd_inc = EpiNow2::covid_incubation_period[1, ]$sd,
  true_mean_gi = parlist$true_mean_GI,
  true_sd_gi = sqrt(parlist$true_var_GI))

## Delays specified as priors, may or may not match truth
testpars$input_lmean_case_delay = testpars$true_log_mean_case_delay
testpars$input_lsd_case_delay = testpars$true_log_sd_case_delay
testpars$input_lmean_death_delay = testpars$true_log_mean_death_delay
testpars$input_lsd_death_delay = testpars$true_log_sd_death_delay
testpars$input_lmean_inc = testpars$true_log_mean_inc
testpars$input_lsd_inc = testpars$true_log_sd_inc
testpars$input_mean_gi = testpars$true_mean_gi
testpars$input_sd_gi = testpars$true_sd_gi
## If output directory does not exist, creat it.
dir_check(testpars$output_folder)
```

## Set up inputs for estimate_infections ---------------------------------------
```{r}
## Set delay distributions for input into EpiNow2 
generation_time <- list(mean = testpars$input_mean_gi, # gamma
                        mean_sd = 1,
                        sd = testpars$input_sd_gi,
                        sd_sd = 1,
                        max = 30)

incubation_period <- list(mean = testpars$input_lmean_inc, # lognormal
                          mean_sd = EpiNow2::covid_incubation_period[1, ]$mean_sd,
                          sd = testpars$input_lsd_inc,
                          sd_sd = EpiNow2::covid_incubation_period[1, ]$sd_sd,
                          max = 30)

obs_rep_delay <- list(mean = testpars$input_lmean_case_delay, ## lognormal, assume mean of 5 days for outpatient testing 
                      mean_sd = 1,   ## (48h from symptom onset + 48h for reporting + report the following day)
                      sd = testpars$input_lsd_case_delay,
                      sd_sd = 1,
                      max = 30)

death_rep_delay <- list(mean = testpars$input_lmean_death_delay, ## lognorma 
                        mean_sd = 1, #For death, assume a mean of 15 days from symptom onset and an sd of about 7d
                        sd = testpars$input_lsd_death_delay,
                        sd_sd = 1,
                        max = 30)
```



## Plot the specified distributions vs the true distributions
```{r echo = F}
{
par(mfrow = c(2,2))
xx = seq(0, 30, by = 0.01)

true_case_dist <- function(x){dlnorm(x, testpars$true_log_mean_case_delay, testpars$true_log_sd_case_delay)}
true_death_dist <- function(x){dlnorm(x, testpars$true_log_mean_death_delay, testpars$true_log_sd_death_delay)} # Additional delay from symptoms -> observation

## Delays to case observation -----
# Input
plot(xx, dlnorm(xx, obs_rep_delay$mean, obs_rep_delay$sd), 
     type = 'l', main = 'assumed delay from onset to case detection', xlab = 'days', ylab = 'dens')
# Truth
lines(xx, true_case_dist(xx), col = 'red', lty = 2)
legend('topright', c('input', 'true'), col = c(1, 'red'), lty = 1)

## Delays to death ------
plot(xx, dlnorm(xx, death_rep_delay$mean, death_rep_delay$sd),
     type = 'l', main = 'assumed delay from onset to death', xlab = 'days', ylab = 'dens')
lines(xx, true_death_dist(xx), col = 'red', lty = 2)
legend('topright', c('input', 'true'), col = c(1, 'red'), lty = 1)

## Generation interval ---
plot(xx, 
     dgamma(xx, shape = with(generation_time, get_shape(mean, sd^2)), 
            rate = with(generation_time, get_rate(mean, sd^2))), 
     type = 'l', main = 'assumed generation time', xlab = 'days', ylab = 'dens')
lines(xx, dgamma(xx, get_shape(testpars$true_mean_gi, testpars$true_sd_gi^2), 
                 get_rate(testpars$true_mean_gi, testpars$true_sd_gi^2)), 
      col = 'red', lty = 2)
legend('topright', c('input', 'true'), col = c(1, 'red'), lty = 1)
## Incubation period
plot(xx, 
     dlnorm(xx, incubation_period$mean, incubation_period$sd), 
     type = 'l', main = 'assumed incubation time', xlab = 'days', ylab = 'dens')
lines(xx, 
      dlnorm(xx, testpars$true_log_mean_inc, testpars$true_log_sd_inc), 
      col = 'red', lty = 2)
legend('topright', c('input', 'true'), col = c(1, 'red'), lty = 1)
}
```



## Generate synthetic observations  -------------------------------------------
```{r}
# Define functions to draw n samples from each delay distribution
r_inc_dist <- function(n){rlnorm(n, testpars$true_log_mean_inc, testpars$true_log_sd_inc)}
# Define delays to observation. Assume lognormal form if an alternate form is not specified on input
r_true_case_delay <- function(n){rlnorm(n, testpars$true_log_mean_case_delay, testpars$true_log_sd_case_delay)}
r_true_death_delay <- function(n){rlnorm(n, testpars$true_log_mean_death_delay, testpars$true_log_sd_death_delay)} # Additional delay from symptoms -> observation

outpatient_delay_dist <- function(nn){r_inc_dist(nn) + r_true_case_delay(nn)}
death_delay_dist <- function(nn){r_inc_dist(nn) + r_true_death_delay(nn)}

source('../00-inf_to_obs.R')
## Wrapper function to get synthetic data in a data frame that includes times of observation
get_obs_ts <- function(mt = NULL, cdelay = outpatient_delay_dist, ddelay = death_delay_dist){ # "Max time" - time at which to end the time series of observations
  if(length(mt) == 0){mt = max(get_sim_df()$time)}
  ## Sample from delays to get times of observation
  get_sim_df() %>%
    merge(
      get_tObs_from_tInf(get_sim_df()$incidence, 
                         get_sim_df()$time, 
                         cdelay, 
                         return_times = T),
      by = 'time', all = TRUE) %>% rename(obs_outpatient = n) %>%
    merge(
      get_tObs_from_tInf(get_sim_df()$incidence, 
                         get_sim_df()$time, 
                         ddelay, 
                         return_times = T),
      by = 'time', all = TRUE) %>% rename(obs_deaths = n) %>%
    as.tbl() -> obs_df
  
  ## Truncate at time mt
  obs_df <- obs_df %>% filter(time <= mt)
  return(obs_df)
}

## Generate synthetic times of observation up to t = 150
obs_df <- get_obs_ts(150) %>%
  mutate(date = Sys.Date()-rev(time)) ## Add calendar dates for input into epinow
```



## Plot synthetic observations
```{r echo = F}
obs_df %>%
  select(date, incidence, contains('obs')) %>%
  mutate(shifted_observed = lead(obs_outpatient, n = round(exp(incubation_period$mean+incubation_period$sd^2/2)+exp(obs_rep_delay$mean+obs_rep_delay$sd^2/2))),
         shifted_deaths = lead(obs_deaths, n = round(exp(incubation_period$mean+incubation_period$sd^2/2)+exp(death_rep_delay$mean+death_rep_delay$sd^2/2)))) %>%
  pivot_longer(-date, names_to = 'data_type', values_to = 'count') %>%
  mutate(data_type = factor(data_type, 
                            levels = c('incidence', 'obs_outpatient', 'obs_deaths', 'shifted_observed', 'shifted_deaths'),
                            labels = c('latent (true) infections\nlater observed as...', 'cases', 'deaths', 'shifted cases', 'shifted deaths'))) %>%
  ggplot() +
  geom_line(aes(x = date, y = count, color = data_type)) +
  ggtitle('Synthetic Data') +
  scale_color_viridis_d(direction = -1) -> obs
obs + guides(color = 'legend') + theme(legend.position = 'right')
```


## Set intputs into estimate_infections
```{r}
get_in_dat <- function(obs_colname, odf = obs_df){
  odf[,c('date', obs_colname)] %>%
    setnames(c('date', 'confirm')) %>%
    as.data.table() %>%
    return()
}

reported_cases = get_in_dat('obs_outpatient')
family = 'negbin'
generation_time = generation_time  ## Generation time priors
incubation_period = incubation_period ## Incubation priors
reporting_delay = obs_rep_delay ## Reporting dealy priors
gp = list(basis_prop = 0.5, boundary_scale = 2)
rt_prior = list(mean = 1, sd = 1) 
adapt_delta = 0.99
max_treedepth = 15
cores = 4
chains = 4 
samples = 2000 
warmup = 500
estimate_rt = TRUE
horizon = 7
verbose = TRUE
return_fit = TRUE
debug = FALSE
prior_smoothing_window = 7
```

## Write functions for diagnosis
```{r}
## Plot the true time series of infections vs the reports and shifted reports
compare_timeseries <- function(ttl = ""){
  
  bind_rows(
    list(reports = reported_cases, 
         shifted_reports = shifted_reported_cases, 
         infections = obs_df %>% mutate(confirm = incidence, day_of_week = NA) %>% select(date, confirm, day_of_week)),
    .id = 'id'
  ) %>%
    ggplot() + 
    geom_line(aes(x = date, y = confirm, color = id)) +
    scale_color_viridis_d() +
    theme(legend.position = c(.2, .8))+
    ggtitle(ttl)
}

## Given an input data frame and column name for cases, find the date of the peak
get_peak_date <- function(df, icol = 'confirm'){
  df %>%
    filter(get(icol) == max(get(icol), na.rm = T)) %>%
    pull(date)
}
```


## Run code from within estimate_infections function
```{r}
# Set up data.table -------------------------------------------------------

suppressMessages(data.table::setDTthreads(threads = cores))

reported_cases <- data.table::setDT(reported_cases)

# Make sure there are no missing dates and order cases --------------------
reported_cases_grid <- data.table::copy(reported_cases)[, .(date = seq(min(date), max(date) + horizon, by = "days"))]

reported_cases <- data.table::merge.data.table(
  reported_cases , reported_cases_grid, 
  by = c("date"), all.y = TRUE)

reported_cases <- reported_cases[is.na(confirm), confirm := 0 ][,.(date = date, confirm)]
reported_cases <- data.table::setorder(reported_cases, date)

## Filter out 0 reported cases from the beginning of the data
reported_cases <- reported_cases[order(date)][,
                                              cum_cases := cumsum(confirm)][cum_cases > 0][, cum_cases := NULL]

# Estimate the mean delay -----------------------------------------------

mean_shift <- as.integer(exp(incubation_period$mean + incubation_period$sd^2/2)+exp(reporting_delay$mean + reporting_delay$sd^2/2))

# Add the mean delay and incubation period on as 0 case days ------------

reported_cases <- data.table::rbindlist(list(
  data.table::data.table(date = seq(min(reported_cases$date) - mean_shift - prior_smoothing_window, 
                                    min(reported_cases$date) - 1, by = "days"),
                         confirm = 0),
  reported_cases
))  
```

## The shift works in the absence of smoothing (frollmean)
```{r}
shifted_reported_cases <- data.table::copy(reported_cases)[,
                                                           confirm := data.table::shift(confirm, n = mean_shift,
                                                                                        type = "lead", 
                                                                                        fill = NA)]

compare_timeseries(ttl = 
                     sprintf('expected mean shift is %.1fd \n actual shift is %sd', 
                             mean_shift,
                             as.Date(get_peak_date(reported_cases))-as.Date(get_peak_date(shifted_reported_cases))
                     )
)
```

## frollmean() seems to shift the peak forward by n days, where n = prior_smoothing_window/2
```{r}
shifted_reported_cases <- data.table::copy(reported_cases)[,
                                                           confirm := data.table::shift(confirm, n = mean_shift,
                                                                                        type = "lead", 
                                                                                        fill = NA)][,
                                                                                                    confirm := data.table::frollmean(confirm, n = prior_smoothing_window, 
                                                                                                                                     align = "right", fill = 0)]

compare_timeseries(ttl = 
                     sprintf('prior smoothing window is %.1fd \n expected mean shift is %.1fd \n actual shift is %sd', 
                             prior_smoothing_window, mean_shift,
                             as.Date(get_peak_date(reported_cases))-as.Date(get_peak_date(shifted_reported_cases))
                     )
)
```
## Resetting the size of the prior smoothing window to 4 decreases the unwanted forward shift
```{r}
prior_smoothing_window = 4
shifted_reported_cases <- data.table::copy(reported_cases)[,
                                                           confirm := data.table::shift(confirm, n = mean_shift,
                                                                                        type = "lead", 
                                                                                        fill = NA)][,
                                                                                                    confirm := data.table::frollmean(confirm, n = prior_smoothing_window, 
                                                                                                                                     align = "right", fill = 0)][,
                                                                                                                                                                 confirm := data.table::fifelse(confirm == 0, 1e-3, confirm)]

compare_timeseries(ttl = 
                     sprintf('prior smoothing window is %.1f \n expected mean shift is %.1fd \n actual shift is %sd', 
                             prior_smoothing_window, mean_shift,
                             as.Date(get_peak_date(reported_cases))-as.Date(get_peak_date(shifted_reported_cases))
                     )
)
```


```